Solving Graph Homomorphism and Subgraph Isomorphism Problems Faster Through Clique Neighbourhood Constraints

### 2021-01-11 paper/paper.tex
More writing, yay

@@

\newtheorem{proposition}{Proposition}

@@

\title{Solving Graph Homomorphism and Subgraph Isomorphism Problems Faster Through Neighbourhood
Maximum Clique Constraints}

@@

Let $G$ and $H$ be graphs. Let $v \in \vertexset(G)$ be a vertex of $G$. The \emph{(open)
neighbourhood} of $v$, written $\neighbourhood_G(v)$, is the set of vertices adjacent to $v$ not
including $v$ itself, whilst the closed neighbourhood of $v$, written $\neighbourhood_G[v]$, is the
neighbourhood plus $v$. The \emph{degree} of a vertex, $\degree_G(v)$, is the cardinality of its open
neighbourhood. Given a vertex set $S \subseteq \vertexset(G)$, the subgraph \emph{induced by} $S$,
written $G[S]$, is the subgraph of $G$ with only the vertices in $S$ together with all the edges
between them. A \emph{clique} is a subgraph where every vertex is adjacent to every other in the
subgraph.

A \emph{homomorphism} from $G$ to $H$ is a function mapping vertices of $G$ to vertices of $H$, such
that adjacent vertices in $G$ are mapped to adjacent vertices in $H$. A homomorphism $h$ is
\emph{locally injective} if for every vertex $w$, the restriction of $h$ to $G[N_G[w]]$ is
injective; if $h$ is injective globally we call it a \emph{(non-induced) subgraph isomorphism}.

@@

?? CSP, hom as CSP

An \emph{implied constraint} is a constraint that can be added to a CSP without altering its set of
solutions. ?? Is it best wording all of this in terms of implied constraints?

\begin{proposition}For the problem of finding a subgraph isomorphism $i$ from a graph $G$ to a graph
    $H$, the following constraints are implied for any vertex $v \in \vertexset(G)$:
    \begin{enumerate}
        \item We cannot map $v$ to a vertex of lower degree, $\degree(v) \le \degree(i(v))$.
        \item Furthermore, ??NDS.
    \end{enumerate}
\end{proposition}

?? Show that these work for locally injective but not homomorphism

\begin{proposition}For the problem of finding a subgraph isomorphism $i$ from a graph $G$ to a graph
    $H$, the following constraints are implied for any pair of vertices $v, w \in \vertexset(G)$:
    \begin{enumerate}
        \item The distance between $v$ and $w$ is at most the distance between $i(v)$ and $i(w)$.
        \item For any pair of vertices $v, w \in \vertexset(G)$, if there are at least $k$ simple
            paths of length exactly $\ell$ between $v$ and $w$, then there must be at least $k$
            simple paths of length exactly $\ell$ between $i(v)$ and $i(w)$.
    \end{enumerate}
\end{proposition}

?? This fourth one is hard. The third one actually works for homomorphisms.

@@

?? Clique, be very careful with target loops. What we want to prove is, if the maximum clique in the
neighbourhood of a pattern vertex $p$ has $n$ vertices, and if the target graph has no loops, then
$p$ cannot be mapped to any target vertex $t$ unless $t$ is contained in a clique of at least $n$
vertices.

?? The point is we can solve clique decision much faster than general homomorphism, and we only need
to do it once per target vertex.

@@

We use the subgraph isomorphism code in the Glasgow Subgraph Solver \cite{DBLP:conf/gg/McCreeshP020}
as a starting point.  We implemented distance filtering for the basic homomorphism problem, using
the existing supplemental graphs framework that the solver uses for subgraph isomorphism to create
pseudo-adjacency constraints for vertices that are distance two away from each other; this was
routine.  We also implemented clique constraints, initially using the Glasgow Subgraph Solver's
existing maximum clique solver to perform domain filtering. Preliminary experiments showed that a
na{\"\i}ve approach, which begins by calculating the maximum clique size for the neighbourhood of
each pattern vertex and each target vertex, would add as much as three minutes of preprocessing time
to some problem instances which could otherwise be solved in a few seconds. We therefore use a more
intelligent approach, which we now describe.

>>>>

\usepackage{cleveref}

@@

\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}

@@

\title{Solving Graph Homomorphism and Subgraph Isomorphism Problems Faster Through Clique Neighbourhood
Constraints}

@@

We begin by introducing notation and terminology, and by giving the theoretical results which
support our implementation.

\paragraph{Graphs.} Let $G$ and $H$ be graphs. Let $v \in \vertexset(G)$ be a vertex of $G$. The
\emph{(open) neighbourhood} of $v$, written $\neighbourhood_G(v)$, is the set of vertices adjacent
to $v$ not including $v$ itself, whilst the closed neighbourhood of $v$, written
$\neighbourhood_G[v]$, is the neighbourhood plus $v$. The \emph{degree} of a vertex, $\degree_G(v)$,
is the cardinality of its open neighbourhood. The \emph{neighbourhood degree sequence} of a vertex
is the sequence consisting of the degrees of its neighbours, in descending order. Given a vertex set
$S \subseteq \vertexset(G)$, the subgraph \emph{induced by} $S$, written $G[S]$, is the subgraph of
$G$ with only the vertices in $S$ together with all the edges between them. A \emph{clique} is a
subgraph where every vertex is adjacent to every other in the subgraph.

\paragraph{Homomorphisms.} A \emph{homomorphism} from $G$ to $H$ is a function mapping vertices of
$G$ to vertices of $H$, such that adjacent vertices in $G$ are mapped to adjacent vertices in $H$. A
homomorphism $h$ is \emph{locally injective} if for every vertex $w$, the restriction of $h$ to
$G[N_G[w]]$ is injective; if $h$ is injective globally we call it a \emph{(non-induced) subgraph
isomorphism}.

@@

\paragraph{Constraint programming.} ?? A constraint programming algorithm.

\paragraph{Degree-based invariants for homomorphisms.} The use of constraint programming techniques
for finding homomorphisms relies upon exploiting various invariants. For example, every recent
constraint programming solver makes use of the following result.

\begin{proposition}\label{proposition:degreends}
    Suppose a subgraph isomorphism $i$ exists from a graph $G$ to a graph $H$. Let
    $v$ be any vertex. Then:
    \begin{enumerate}
        \item The function cannot map $v$ to a vertex of lower degree, $\degree(v) \le \degree(i(v))$.
        \item Furthermore, $v$ cannot be mapped to any vertex whose neighbourhood degree sequence is
            not pointwise greater than or equal to its own
            \cite{DBLP:journals/constraints/ZampelliDS10}.
    \end{enumerate}
\end{proposition}

The simplest way for a constraint programming algorithm to use these results is as unary
constraints, which are propagated when domains are initialised, before search begins. For each
domain representing a pattern vertex, any value representing target vertex whose degree or
neighbourhood degree sequence is too low may immediately be rejected.

These invariants do \emph{not} hold for homomorphisms in general. As a simple counter-example, a
homomorphism may map a star graph (one central vertex of high degree, which is adjacent to many
other vertices, each of which has no other neighbours) onto a single edge. However, it is
straightforward to verify the following.

\begin{proposition}Both invariants in \cref{proposition:degreends} are also valid for locally
injective homomorphisms.\end{proposition}

\paragraph{Path-based invariants for homomorphisms.} Constraint programming solvers can also exploit
invariants that are based upon paths. This is done by automatically adding \emph{implied}
constraints to the problem that are implied by the original model, but that will give stronger
propagation power.

\begin{proposition}For the problem of finding a subgraph isomorphism $i$ from a graph $G$ to a graph
    $H$, the following constraints are implied for any pair of vertices $v, w \in \vertexset(G)$:
    \begin{enumerate}
        \item The distance between $v$ and $w$ is at most the distance between $i(v)$ and $i(w)$
            \cite{DBLP:conf/cp/AudemardLMGP14}.
        \item If there are at least $k$ simple
            paths of length exactly $\ell$ between $v$ and $w$, then there must be at least $k$
            simple paths of length exactly $\ell$ between $i(v)$ and $i(w)$
            \cite{DBLP:conf/cp/McCreeshP15}.
    \end{enumerate}
\end{proposition}

It is easy to verify that the first of these two properties is valid for \emph{any} homomorphism.
The second property does not hold for homomorphisms in general: for example, a pair of vertices
connected by three paths of length two may be mapped onto a pair of vertices connected by a single
path of length two. Local injectivity is trickier. The second property does not hold in full
generality, because, for example, a path of length three may be mapped onto a triangle whilst
respecting local injectivity. However, a weaker result holds:

\begin{proposition}For the problem of finding a locally injective graph homomorphism $i$ from a
    graph $G$ to a graph $H$, for any pair of vertices $v, w \in \vertexset(G)$, if there are at
    least $k$ simple paths of length exactly two between $v$ and $w$, then there must be at least
    $k$ simple paths of length exactly two between $i(v)$ and $i(w)$.
\end{proposition}

\begin{proof}Let $\{ x_1, \ldots, x_n \}$ be the intermediate vertices on the paths of length two
between $v$ and $w$. Observe that each $x_j$ is in the neighbourhood of $v$, and so must be mapped
    to different vertices. Thus each sequence $(i(v), i(x_i), i(w))$ gives a distinct simple path of
length two between $i(v)$ and $i(w)$ as required.\end{proof}

@@

\paragraph{Cliques and homomorphisms.} Finally, we come to the main result of this paper. Observe
that until now, we have not seen any unary constraints which are valid for the general homomorphism
problem. All existing constraint programming approaches for the subgraph isomorphism problem begin
by branching on the variable which has the smallest domain---typically this will be a domain for a
pattern vertex of high degree, or which has many high degree neighbours. Furthermore, once one such
guessed assignment has been made, adjacency propagation means many other domains will be
substantially reduced in size, making subsequent branching choices simpler.  However, for
homomorphism problems, if we cannot find any implied unary constraints then every domain will
initially be of the same size, which will make it much harder for a solver to know where to start.
This motivates the following observation.

\begin{proposition}\label{proposition:clique}
    Suppose we are finding a graph homomorphism $i$ from a graph $G$ to a graph $H$ where $H$ does
    not contain any loops. Suppose there is a $k$ vertex clique in $G$. Then the image of this
    clique under $i$ must also be a $k$ vertex clique.
\end{proposition}

\begin{proof}
    Let $j$ and $k$ be two distinct vertices of the clique in $G$. By definition of a homomorphism,
    $i(j)$ and $i(k)$ must be adjacent. And, because $H$ does not contain loops, $i(j)$ cannot equal
    $i(k)$.
\end{proof}

\begin{corollary}\Cref{proposition:clique} holds for locally injective graph homomorphisms and for
    subgraph isomorphisms even if loops are present in the second graph.
\end{corollary}

We will discuss how this is useful in practice in the following section.

@@

Having determined which commonly-used subgraph isomorphism invariants do and do not hold for other
forms of homomorphism, and having discovered a new clique-based invariant, we will now look at how
these invariants may be used in practice. Our starting point is the Glasgow Subgraph Solver
\cite{DBLP:conf/gg/McCreeshP020}, due to it being the current single strongest solver for the
subgraph isomorphism problem.

??

We implemented distance filtering for the basic homomorphism problem, using
the existing supplemental graphs framework that the solver uses for subgraph isomorphism to create
pseudo-adjacency constraints for vertices that are distance two away from each other; this was
routine.

We also implemented clique constraints, initially using the Glasgow Subgraph Solver's existing
maximum clique solver to perform domain filtering. Preliminary experiments showed that a na{\"\i}ve
approach, which begins by calculating the maximum clique size for the neighbourhood of each pattern
vertex and each target vertex, would add as much as three minutes of preprocessing time to some
problem instances which could otherwise be solved in a few seconds. We therefore use a more
intelligent approach, which we now describe.
